Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-02-28 manuscript.tex
Restructured introduction
Still in the form of  loose phases. Need to tie them together.

We use the optimized formula of \citet{Grombein2013} with Cartesian integral
kernels.

@@

The traditional spherical integral kernels suffer from a singularity at the
poles \citep{Heck2007, Wild-Pfeiffer2008}.

@@

The Taylor series expansion approach produces accurate results at low
latitudes.

The Taylor series expansion presents a decrease in accuracy towards the polar
regions.

The tesseroids have an approximately rectangular surface area at low latitudes
but collapse into an approximately triangular shape at the poles.

\citet{Grombein2013} use a near-zone separation to mitigate the increased error
in high latitudes.

@@

We use the Gauss-Legendre Quadrature (GLQ) to numerically integrate the Cartesian
kernels instead of the Taylor series expansion \citep{Asgharzadeh2007}.

The GLQ integration consists of approximating the volume integral by a weighted sum of
the effect of point masses.

The point masses are distributed according to the roots of Legendre polynomials
\citep{Hildebrand1987}.

A disadvantage of the GLQ is the increased computation time as the number of
point masses increases.

There is a trade-off between accuracy and computation time.

\citet{Ku1977} suggests that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.

\citet{Ku1977} proposes an empirical criteria that the distance between adjacent
point masses should be less than the distance to the computation point.

This empirical criteria was used by \citet{Asgharzadeh2007} as a way to
choose the number of point masses used in the GLQ integration.

They suggested using this criteria for gravitational attraction as well as
the gravity gradient tensor (or Marussi tensor) of a tesseroid.

\citet{Ku1977} presented results for the vertical component of the
gravitational acceleration ($g_z$) caused by a right rectangular prism.

To our knowledge, there has been no investigation if the empirical criteria of
\citet{Ku1977} is valid for the second derivatives of the
gravitational potential.

There has also been no attempt to quantify the error committed in the GLQ
integration of gravity gradients when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an automatic algorithm to enforce the criteria of
\citet{Ku1977}.

Their algorithm keeps the number of points masses per tesseroid fixed.

They then check the ratio between the minimum distance to the computation point
and the largest dimension of the tesseroid.

If the ratio below a specified threshold value, the tesseroid is divided into
smaller tesseroids.

The GLQ integration is then performed for each of the smaller tesseroids
using the fixed number of point masses.

The advantage of the adaptive discretization of \citet{Li2011} over simply
increasing the number of points masses in the GLQ is that the
amount of point masses will be greater only close to the computation point.

This article describes the software \textit{Tesseroids} and the implementation
of the adaptive discretization algorithm.

@@

We will attempt to quantify the error committed in the computation of the
gravity gradients using the GLQ with adaptive discretization.

We will establish an empirical relationship between the integration error and
the ratio between the distance to the computation point and the dimensions of
the tesseroid.

This will allow one to choose the appropriate amount of discretization to
achieve the desired accuracy.

>>>>

Satellite missions dedicated to measuring the Earths gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.

These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.

Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.

A common approach to achieve this is
to discretize the Earth into tesseroids (spherical prisms)
instead of rectangular prisms.

However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.

Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.

Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.

This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.

The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.

A disadvantage is the increased computation time as the number of
point masses increases.

Thus, there is a trade-off between accuracy and computation time, a common
theme in numerical methods.

\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.

The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.

However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the Gauss-Legendre
Quadrature in gravity forward modeling.

\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.

The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.

Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.

\citet{Asgharzadeh2007} proposed using this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.

To our knowledge, there has been no investigation if the empirical criteria of
\citet{Ku1977} is valid for gravity gradient components or for tesseroids.

There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.

Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.

This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.

The criterion used decide if a division is necessary is
that the minimum distance to the computation point must be larger than
the largest dimension of the tesseroid.

Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.

The advantage of this adaptive discretization over
increasing the number of points masses is that the
total total distribution of point masses will be greater
only close to the computation point.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.

@@

@@

\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.

@@

We have implemented a modified version of the adaptive discretion of
\citet{Li2011} into the open-source software package \textit{Tesseroids}.

The software uses the Cartesian formula of \citet{Grombein2013} for improved
performance and robustness.

@@

This article describes the software \textit{Tesseroids} and the implementation
of our modified adaptive discretization algorithm.

We also present a numerical investigation of the error committed
in the computations.

These results allow us calibrate the adaptive discretization algorithm
separately for the gravitational potential, gravitational acceleration,
as well as the gravity gradient tensor components.
